---
title: "Penetrometer"
author: "Rebecca Martinez"
date: "11-18-2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,     # hide code
  warning = FALSE,  # hide warnings
  message = FALSE,  # hide messages
  error = FALSE     # hide errors
)

#install.packages("tidyverse")
#install.packages("gt")
#install.packages("flextable")
#intall.packages("janitor")
#install.packages("cowplot")
library(tidyverse)
library(janitor) #clean data 
library(gt)
library(flextable)
library(cowplot) # plots into objects, insert objects to make grid

```

```{r data}
penetromer_raw <- read_csv("data_raw/wide_penetrometer.csv") 

penetromer_clean <- clean_names(penetromer_raw) |> 
  select(-pd_blows_m, -notes)

```

## Methods and Measurments


The drop cone method measures the depth of penetration resulting from a cone of fixed mass being dropped from a standard height. The hammer-type penetrometers use a slide hammer of fixed mass and drop height to apply consistent kinetic energy with each blow. Either the number of blows required to penetrate a specified depth, or the depth of penetration per blow are measured in this method.

Soil penetration resistance \(R_s\) is defined as the work done by the soil divided by the distance the penetrometer travels:

\[
R_s = \frac{W_s}{P_d}
\]

where \(R_s\) is the soil resistance (N), \(W_s\) is the work done by the soil (J), and \(P_d\) is the distance the penetrometer travels (m).

The kinetic energy of a hammer of mass \(m = 4.58 \text{ kg}\) falling a distance \(x = 0.45 \text{ m}\) is:

\[
v = \sqrt{v_0^2 + 2 a x} = \sqrt{0 + 2 (9.8)(0.45)} = 2.97 \text{ m/s}
\]

\[
KE = \frac{1}{2} m v^2 = \frac{1}{2} (4.58)(2.97)^2 = 20.20 \text{ J}
\]

Then the soil resistance is calculated as:

\[
R_s = \frac{W_s}{P_d} = \frac{KE}{P_d}
\]




```{r}

sqrt(2*(9.8*.45))

.5*(4.58*(2.97^2))



```

