---
title: "Penetrometer"
author: "Rebecca Martinez"
date: "11-18-2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Methods and Measurments


The drop cone method measures the depth of penetration resulting from a cone of fixed mass being dropped from a standard height. These have been successfully used to measure shear strength in soils (10; 14). The hammer-type penetrometers use a slide hammer of fixed mass and drop height to apply consistent kinetic energy with each blow. Either the number of blows required to penetrate a specified depth, or the depth of penetration per blow are measured in this method.

Soil penetration resistance \(R_s\) is defined as the work done by the soil divided by the distance the penetrometer travels:

\[
R_s = \frac{W_s}{P_d}
\]

where \(R_s\) is the soil resistance (N), \(W_s\) is the work done by the soil (J), and \(P_d\) is the distance the penetrometer travels (m).

The kinetic energy of a hammer of mass \(m = 2 \text{ kg}\) falling a distance \(x = 0.4 \text{ m}\) is:

\[
v = \sqrt{v_0^2 + 2 a x} = \sqrt{0 + 2 (9.8)(0.4)} = 2.8 \text{ m/s}
\]

\[
KE = \frac{1}{2} m v^2 = \frac{1}{2} (2)(2.8)^2 = 7.84 \text{ J}
\]

Then the soil resistance is calculated as:

\[
R_s = \frac{W_s}{P_d} = \frac{KE}{P_d}
\]



```{r}





```

